---
title: "Stat231: Lab 9 - Unsupervised Learning" 
author: "Your Name Here"
date: "Sprring 2021"
always_allow_html: yes
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggrepel)
library(GGally)

sessionInfo()

knitr::opts_chunk$set(eval = TRUE
                      , message = FALSE
                      , warning = FALSE
                      , fig.width = 9
                      , fig.height = 6)
```


# K-means clustering 

## SAT Scores

Let's begin with a simple example of clustering some colleges and universities based on characteristics of an incoming class of first year students.  The variables we have are:

- `STABBR`: State 
- `CITY`: City
- `INSTNM`: Institution
- `SATMT25`: 25th percentile SAT MATH score
- `SATVR25`: 25th percentile SAT Verbal score
- `ADM_RATE`: Admission rate
- `SAT_AVG`: Average SAT score
- `GRAD_DEBT_MDN`: Median debt at graduation ($)
- `PCIP27`: % of graduates majoring in Mathematics & Statistics
- `COSTT4_A`: Average cost of attendance 

To simplify our look at this data, we will only look at a random sample of schools in Massachusetts, and we will focus on SAT scores for right now.  

```{r}
colleges <- read_csv("colleges_subset.csv")

dim(colleges)
head(colleges)

set.seed(20201103)
ma_samp <- colleges %>% 
  mutate(SATMT = as.numeric(SATMT25)
         , SATVR = as.numeric(SATVR25)) %>%
  # only keep schools in MA with non-missing SAT scores
  filter(STABBR %in% c("MA") & !is.na(SATMT) & !is.na(SATVR)) %>%
  select(STABBR, CITY, INSTNM, SATMT, SATVR) %>%
  # select a random sample of 15 schools
  sample_n(15)

ma_samp$INSTNM
```

First, let's just look at a scatterplot of Math SAT (25th percentile) vs Verbal SAT (25th percentile):

```{r}
ggplot(data = ma_samp, aes(x = SATMT, y = SATVR)) +
  geom_point() + 
  geom_text_repel(aes(label=INSTNM), size = 3) +
  labs(x = "Math SAT 25th Percentile", y = "Verbal SAT 25th Percentile")
```

Which schools are most similar in terms of these two variables?

> ANSWER:  Framingham State and Fitchburg State are the most similar in terms of these two variables, but more generally there seem to be two distinct clusters of schools, one in the upper right corner and the other in the lower left.

Do we need to standardize the variables in this case? Why or why not?

> ANSWER: We don't need to standardize the variables in this case because they are already both measured using the same scale (score out of 800).

Let's use k-means to find two clusters:

```{r}
set.seed(23)
vars <- c("SATMT", "SATVR")
km2_out <- kmeans(ma_samp[,vars], centers=2, nstart=20)

# vector of cluster assignments
km2_out$cluster

# the centroids for the fit
km2_out$centers

# add cluster assignment to the data frame
ma_samp_clust2 <- ma_samp %>%
  mutate(clust2 = as.character(km2_out$cluster))

# visualize the cluster assignments and centroids
ggplot(data = ma_samp_clust2, aes(x = SATMT, y = SATVR)) + 
  geom_point(aes(color = clust2)) +
  geom_text_repel(aes(label=INSTNM, color = clust2), size = 3) +
  coord_fixed() +
  geom_point(data = as.data.frame(km2_out$centers)
             , aes(x = SATMT, y = SATVR)
             , pch = "X"
             , size = 4) +
  labs(x = "Math SAT 25th Percentile"
       , y = "Verbal SAT 25th Percentile" 
       , color = "Cluster Assignment")
```

What if we used 5 clusters?  Repeat the k-means clustering analysis and visualization above, but with 5 clusters instead of 2.  

```{r}
set.seed(23)
vars <- c("SATMT", "SATVR")
km5_out <- kmeans(ma_samp[,vars], centers=5, nstart=20)

# vector of cluster assignments
km5_out$cluster

# the centroids for the fit
km5_out$centers

# add cluster assignment to the data frame
ma_samp_clust5 <- ma_samp %>%
  mutate(clust5 = as.character(km5_out$cluster))

# visualize the cluster assignments and centroids
ggplot(data = ma_samp_clust5, aes(x = SATMT, y = SATVR)) + 
  geom_point(aes(color = clust5)) +
  geom_text_repel(aes(label=INSTNM, color = clust5), size = 3) +
  coord_fixed() +
  geom_point(data = as.data.frame(km5_out$centers)
             , aes(x = SATMT, y = SATVR)
             , pch = "X"
             , size = 4) +
  labs(x = "Math SAT 25th Percentile"
       , y = "Verbal SAT 25th Percentile" 
       , color = "Cluster Assignment")
```

Which solution will have *smaller* total within-cluster variation?  Why?  Check your answer using the code below.

> ANSWER: The solution with five clusters will have a smaller total within-cluster variation, because when there are more clusters these clusters can afford to be smaller, giving each individual point more say in where the center is, and resulting in the points in each cluster being closer to their respective center.

```{r}
km2_out$tot.withinss
km5_out$tot.withinss
```


### Plotting Total Within Cluster Variation vs. K

```{r}
fig <- matrix(NA, nrow=10, ncol=2)

set.seed(75)
for (i in 1:10){
  fig[i,1] <- i
  fig[i,2] <- kmeans(ma_samp[,vars]
                    , centers=i
                    , nstart=20)$tot.withinss
}

# elbow plot
ggplot(data = as.data.frame(fig), aes(x = V1, y = V2)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks=c(1:10)) +
  labs(x = "K", y = expression("Total W"[k]))

# start the x-axis at k=2
ggplot(data = as.data.frame(fig[2:10,]), aes(x = V1, y = V2)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks=c(1:10)) +
  labs(x = "K", y = expression("Total W"[k]))
```

### Standardization

Let's try clustering colleges by more than two variables.  This time, we'll consider the following set of characteristics:

- Admission rate
- SAT average
- Median debt at graduation
- % of graduates majoring in Mathematics & Statistics
- Average cost of attendance

This will be harder to visualize, but since we only have 15 observations, we can list out which colleges end up in each cluster.

```{r}
# add the additional variables needed to our sample dataset 
ma_samp2 <- colleges %>%
  select(INSTNM, ADM_RATE, SAT_AVG
         , GRAD_DEBT_MDN, PCIP27, COSTT4_A) %>%
  # use right_join to only keep the colleges in our original sample
  right_join(ma_samp, by = "INSTNM") %>%
  # these variables are character but should be numeric
  mutate_at(c("ADM_RATE", "SAT_AVG", "GRAD_DEBT_MDN"
              , "COSTT4_A", "PCIP27"), ~as.numeric(.)) %>%
  # standardize the numeric variables using the scale() function
  # scale() subtracts the mean and divides by the standard deviation
  mutate_if(is.numeric, funs(`std`=scale(.) %>% as.vector())) %>% 
  janitor::clean_names() %>% 
  rename(avg_cost = costt4_a
         , avg_cost_std = costt4_a_std
         , prop_math = pcip27
         , prop_math_std = pcip27_std) %>%
  # drop 3 extra Wentworth rows with missing values
  # note that kmeans function won't work if there are NAs
  drop_na()

glimpse(ma_samp2)
  
# k=3 
set.seed(23)
## on standardized variables
vars_std <- c("adm_rate_std", "sat_avg_std", "grad_debt_mdn_std"
              , "avg_cost_std", "prop_math_std")
km3_out_std <- kmeans(ma_samp2[,vars_std]
                        , centers=3
                        , nstart=20)

## on unstandardized variables (for comparison)
vars_unstd <- c("adm_rate", "sat_avg", "grad_debt_mdn"
                , "avg_cost", "prop_math")
km3_out_unstd <- kmeans(ma_samp2[,vars_unstd]
                           , centers=3
                           , nstart=20)


# add cluster assignments to the data frame
ma_samp2_clust3 <- ma_samp2 %>%
  mutate(clust3_std = as.character(km3_out_std$cluster)
         ,clust3_unstd = as.character(km3_out_unstd$cluster))
```

What schools are clustered together when using the standardized variables?  Are the clusters different if the unstandardized variables are used?

> ANSWER: When standard variables are used, Amherst College is in a cluster all its own, and Fitchburg and Framingham State cluster with Simmons, Gordon, and Wentworth. If unstandardized variables are used, Amherst is clustered with the cluster 1 schools from the standardized variables, and cluster 3 from the standardized separates into a cluster for state schools and a cluster for Simmons, Gordon, and Wentworth.

```{r}
# clusters based on standardized vars
ma_samp2_clust3 %>%
  select(instnm, clust3_std) %>%
  arrange(clust3_std)

# clusters based on unstandardized vars
ma_samp2_clust3 %>%
  select(instnm, clust3_unstd) %>%
  arrange(clust3_unstd)
```

### Extracting meaning

Clustering methods don't just determine clusters based on individual variable values, but how these variables combine with all other variables (e.g. clusters in $R^5$ space for this example). We can also look at a 2D view of the data using a scatterplot matrix. We can use the ggplot "add on" package called `GGally` (loaded at the top of this program) to do this:

```{r}
#head(ma_samp2_clust3)
#vars_unstd
#vars_std

# unstandardized grid plot
ggpairs(data = ma_samp2_clust3
        , aes(color = clust3_unstd) 
        , columns = vars_unstd)

# standardized grid plot
ggpairs(data = ma_samp2_clust3
        , aes(color = clust3_std) 
        , columns = vars_std
        , upper = list(continuous = "blank"))
```

Consider both the density plots and the scatterplots.  What are the defining characteristics of each cluster?

> ANSWER: In the unstandardized version, cluster 1 is defined by low admission rate, high SAT scores, and high average cost. In the standardized version, it is defined similarly, but Amherst is not included because of its unusually low graduate debt and unusually high proportion of math and stat majors. In the unstandardized version, cluster 2 is defined by high admission rates, low average cost, and low proportion of math and stat majors. In the standardized version these schools become a part of cluster 3, which is associated with the same factors, as well as high graduate debt.


\newpage
## Your Turn!  Customer segmentation

Suppose you're working as a data scientist for a credit card company.  The company wants to divide their customers into groups for targeted marketing.  That is, your tasked with grouping the credit card holders based on their credit card behavior; then, the marketing team at the company will use the information you provide them to help inform their marketing stategy.

You're given a dataset with information on 8,950 credit card holders, with the following variables:

- `CUST_ID`: Identification of Credit Card holder
- `BALANCE`: Balance amount left in their account to make purchases 
- `BALANCE_FREQUENCY`: How frequently the Balance is updated, score between 0 and 1 (1 = frequently updated, 0 = not frequently updated)
- `PURCHASES`: Amount of purchases made from account
- `PURCHASES_FREQUENCY`: How frequently the Purchases are being made, score between 0 and 1 (1 = frequently purchased, 0 = not frequently purchased)
- `ONEOFF_PURCHASES`: Maximum purchase amount done in one-go
- `ONEOFFPURCHASESFREQUENCY`: How frequently purchases are happening in one-go (1 = frequently purchased, 0 = not frequently purchased)
- `PRC_FULL_PAYMENT`: Percent of full payment paid by user 
- `INSTALLMENTS_PURCHASES`: Amount of purchase done in installment
- `PURCHASES_INSTALLMENTS_FREQUENCY`: How frequently purchases in installments are being done (1 = frequently done, 0 = not frequently done)- `CASH_ADVANCE`: Cash in advance given by the user
- `CASH_ADVANCE_FREQUENCY`: How frequently the cash in advance being paid
- `CASH_ADVANCE_TRX`: Number of Transactions made with "Cash in Advanced"
- `PURCHASES_TRX`: Number of purchase transactions made
- `CREDIT_LIMIT`: Limit of Credit Card for user
- `PAYMENTS`: Amount of Payment done by user
- `MINIMUM_PAYMENTS`: Minimum amount of payments made by user
- `PRC_FULL_PAYMENT`: Percent of full payment paid by user
- `TENURE`: Tenure of credit card service for user

```{r}
cc <- read_csv("CC GENERAL.csv") 

glimpse(cc)
summary(cc)

#ggplot(data=cc, aes(x=TENURE)) + geom_histogram()
```

1. Apply k-means clustering to identify 3 clusters.  Don't forget to remove any rows with missing values, and to standardize the variables first.  How many customers are in each cluster?  

> ANSWER: There are 1197 customers in cluster 1, 1533 in cluster 2, and 5906 in cluster 3.

```{r}
cc_clusts <- cc %>%
  mutate_if(is.numeric, funs(`std`=scale(.) %>% as.vector())) %>% 
  janitor::clean_names() %>% 
  drop_na()
km_out <- kmeans(cc_clusts[, c(19 : 35)], centers=3, nstart=20)
cc_totals <- cc_clusts %>%
  mutate(clust = as.character(km_out$cluster)) %>%
  group_by(clust) %>%
    summarise(total = n())
```

2. Computer the centroids for each cluster.  Can you identify any distinguishing characteristics about the clusters from these centroid values? 

> ANSWER: Cluster 1 is characterized by high purchasing, Cluster 2 is characterized by high balance and cash advance, and Cluster 3 is characterized by low balance, purchases, and payment.

```{r}
km_out$centers
```

3. In 1-3 sentences, explain how you would expect the results to be different had you forgotten to standardize the variables prior to clustering.

> ANSWER: If I had forgotten to standardize, I would expect the large variables like balance, credit limit, and minimum payment to have a much greater influence on the cluster assignments than smaller variables such as purchases frequency.

4. Identify some primary defining characteristics of each of the three clusters.  Come up with a short name for each cluster based on their defining characteristics. 

Note:  There are so many variables that using `ggpairs` to visualize them all at once produces a figure that isn't legible.  I recommend running `ggpairs` separately for 3-4 variables at a time and/or considering the centroids of the clusters (as you computed above in #2).  

> ANSWER: Cluster 3 I will call the Meager Cluster based on their relatively small balances. Cluster 1 I will call the Spender Cluster because of their frequency of purchasing, oneoff purchasing, and installment purchasing. Cluster 2 I will call the Scrooge Cluster, since they have a lot of money in their balance and update it frequently but don't buy much.

```{r}
bal_vars <- c("balance_std", "balance_frequency_std", "prc_full_payment_std")
purchase_vars <- c("purchases_frequency_std", "oneoff_purchases_frequency_std", "purchases_installments_frequency_std")
ggpairs(data = cc_clust
        , aes(color = clust) 
        , columns = bal_vars)
ggpairs(data = cc_clust
        , aes(color = clust) 
        , columns = purchase_vars)
```

5. Create an "elbow" plot to help identify an appropriate number of clusters to create in this analysis.  How many clusters seem reasonable such that there's enough of a decrease in the total within cluster variability to warrant that many clusters but not too many clusters to complicate the analysis.  How might you proceed?

> ANSWER: The greatest decrease occurs between one cluster and two clusters, but I think this is always the case, and other than that the most distinct change in slope occurs between 3 and 5 clusters, after which the graph becomes quite flat. With this in mind, I would advocate using 4 clusters for this analysis.

```{r}
cc_fig <- matrix(NA, nrow=10, ncol=2)

set.seed(75)
for (i in 1:10){
  cc_fig[i,1] <- i
  cc_fig[i,2] <- kmeans(cc_clusts[,c(19 : 35)]
                    , centers=i
                    , nstart=20)$tot.withinss
}

# elbow plot
ggplot(data = as.data.frame(cc_fig), aes(x = V1, y = V2)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks=c(1:10)) +
  labs(x = "K", y = expression("Total W"[k]))

# start the x-axis at k=2
ggplot(data = as.data.frame(cc_fig[2:10,]), aes(x = V1, y = V2)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks=c(1:10)) +
  labs(x = "K", y = expression("Total W"[k]))

```


# References

College data is from: https://collegescorecard.ed.gov/data/
